---
title: "Fast Comprehensive Dataset Analysis: North America vs Global"
subtitle: "Complete analysis with efficient processing"
author: "Sebastian Dohne"
date: "June 2025"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Executive Summary

**Should you use North American observations or the full global dataset for your wheat yield ML analysis?**

This analysis examines **ALL VARIABLES** efficiently and provides comprehensive comparisons.

```{r libraries}
# Load essential libraries only
library(dplyr, warn.conflicts = FALSE)
```

```{r data-loading}
# Load and examine the complete dataset
cat("=== LOADING COMPLETE DATASET ===\n")
data <- read.csv("../../../../data/Data/chatgptdata/23-05-mergedwheatdatafixed.csv", fileEncoding = "utf-8")


cat("Original dataset dimensions:", nrow(data), "rows x", ncol(data), "columns\n")

# Display ALL column names
cat("\n=== COMPLETE VARIABLE INVENTORY ===\n")
all_vars <- names(data)
cat("Total variables:", length(all_vars), "\n\n")

# Categorize variables by type
yield_vars <- grep("yield|Yield", all_vars, value = TRUE, ignore.case = TRUE)
soil_vars <- grep("sand|silt|clay|pH|soil|organic|carbon", all_vars, value = TRUE, ignore.case = TRUE)
climate_vars <- grep("temp|prc|precipitation|temperature|climate", all_vars, value = TRUE, ignore.case = TRUE)
management_vars <- grep("rate|fertilizer|N\\.|P\\.|K\\.", all_vars, value = TRUE, ignore.case = TRUE)
location_vars <- grep("country|longitude|latitude|elevation|AEZ|region", all_vars, value = TRUE, ignore.case = TRUE)
temporal_vars <- grep("year|date|time|season", all_vars, value = TRUE, ignore.case = TRUE)
other_vars <- setdiff(all_vars, c(yield_vars, soil_vars, climate_vars, management_vars, location_vars, temporal_vars))

cat("VARIABLE CATEGORIES:\n")
cat("Yield variables (", length(yield_vars), "):", paste(yield_vars, collapse = ", "), "\n")
cat("Soil variables (", length(soil_vars), "):", paste(soil_vars, collapse = ", "), "\n") 
cat("Climate variables (", length(climate_vars), "):", paste(climate_vars, collapse = ", "), "\n")
cat("Management variables (", length(management_vars), "):", paste(management_vars, collapse = ", "), "\n")
cat("Location variables (", length(location_vars), "):", paste(location_vars, collapse = ", "), "\n")
cat("Temporal variables (", length(temporal_vars), "):", paste(temporal_vars, collapse = ", "), "\n")
cat("Other variables (", length(other_vars), "):", paste(other_vars, collapse = ", "), "\n\n")
```

```{r data-cleaning}
# Fast data cleaning
cat("=== FAST DATA CLEANING ===\n")

# Clean yield data
if("Grain.yield..tons.ha.1." %in% names(data)) {
  valid_yield <- !is.na(data$Grain.yield..tons.ha.1.) & 
                 data$Grain.yield..tons.ha.1. > 0 & 
                 data$Grain.yield..tons.ha.1. < 50
  
  clean_data <- data[valid_yield, ]
  cat("Valid yield observations:", nrow(clean_data), "out of", nrow(data), "\n")
} else {
  clean_data <- data
}

# Clean coordinates if available
if(all(c("Conversion.for.longitude", "Conversion.for.latitude") %in% names(clean_data))) {
  valid_coords <- !is.na(clean_data$Conversion.for.longitude) & 
                  !is.na(clean_data$Conversion.for.latitude)
  clean_data <- clean_data[valid_coords, ]
  cat("Valid coordinate observations:", nrow(clean_data), "\n")
}

# Add derived variables
clean_data$yield_value <- as.numeric(clean_data$Grain.yield..tons.ha.1.)

cat("Final clean dataset:", nrow(clean_data), "observations\n\n")
```

```{r dataset-creation}
# Create comparison datasets
cat("=== CREATING COMPARISON DATASETS ===\n")

# North America
na_countries <- c("United States of America", "Canada", "Mexico", "USA", "US", "United States")
north_america <- clean_data[clean_data$Country %in% na_countries, ]

# Global dataset
global_dataset <- clean_data

cat("Dataset sizes:\n")
cat("North America only:", nrow(north_america), "observations\n")
cat("Global (including NA):", nrow(global_dataset), "observations\n")
cat("NA as % of Global:", round(nrow(north_america) / nrow(global_dataset) * 100, 1), "%\n\n")

# Country distribution
cat("Geographic coverage:\n")
cat("Countries in North America dataset:", length(unique(north_america$Country)), "\n")
cat("Countries in Global dataset:", length(unique(global_dataset$Country)), "\n\n")
```

## 1. Complete Variable Analysis

```{r variable-analysis}
cat("=== COMPLETE VARIABLE AVAILABILITY ANALYSIS ===\n")

# Fast variable completeness analysis
analyze_completeness_fast <- function(dataset, dataset_name) {
  cat(dataset_name, "Variable Analysis:\n")
  
  # Calculate completeness for ALL variables quickly
  completeness <- sapply(dataset, function(x) {
    non_missing <- sum(!is.na(x) & x != "" & x != "NA")
    round(non_missing / nrow(dataset) * 100, 1)
  })
  
  # Quality categories
  excellent <- sum(completeness >= 95)
  good <- sum(completeness >= 80 & completeness < 95)
  fair <- sum(completeness >= 50 & completeness < 80)
  poor <- sum(completeness < 50)
  
  cat("Variable quality distribution:\n")
  cat("  Excellent (≥95%):", excellent, "variables\n")
  cat("  Good (80-94%):", good, "variables\n")
  cat("  Fair (50-79%):", fair, "variables\n")
  cat("  Poor (<50%):", poor, "variables\n")
  
  mean_completeness <- round(mean(completeness), 1)
  cat("  Mean completeness:", mean_completeness, "%\n\n")
  
  return(list(
    completeness = completeness,
    mean_completeness = mean_completeness,
    excellent = excellent,
    good = good,
    fair = fair,
    poor = poor
  ))
}

# Analyze both datasets
na_analysis <- analyze_completeness_fast(north_america, "NORTH AMERICA")
global_analysis <- analyze_completeness_fast(global_dataset, "GLOBAL")

# Compare overall quality
cat("OVERALL QUALITY COMPARISON:\n")
cat("North America mean completeness:", na_analysis$mean_completeness, "%\n")
cat("Global mean completeness:", global_analysis$mean_completeness, "%\n")
improvement <- na_analysis$mean_completeness - global_analysis$mean_completeness
cat("North America improvement:", sprintf("%+.1f", improvement), "percentage points\n\n")

# Top and bottom variables comparison
na_comp <- na_analysis$completeness
global_comp <- global_analysis$completeness
var_comparison <- na_comp - global_comp[names(na_comp)]

cat("Variables with biggest improvements in North America:\n")
top_improvements <- sort(var_comparison, decreasing = TRUE)[1:10]
for(i in 1:length(top_improvements)) {
  if(top_improvements[i] > 0) {
    cat("  ", names(top_improvements)[i], ": +", round(top_improvements[i], 1), "% points\n")
  }
}

cat("\nVariables with biggest deteriorations in North America:\n")
bottom_improvements <- sort(var_comparison, decreasing = FALSE)[1:10]
for(i in 1:length(bottom_improvements)) {
  if(bottom_improvements[i] < 0) {
    cat("  ", names(bottom_improvements)[i], ": ", round(bottom_improvements[i], 1), "% points\n")
  }
}
```

## 2. Effective Sample Sizes

```{r sample-size-analysis}
cat("=== EFFECTIVE SAMPLE SIZE ANALYSIS ===\n")

# Define variable sets
variable_sets <- list(
  minimal = c("Grain.yield..tons.ha.1.", "Sand", "Silt", "Clay"),
  basic_soil = c("Grain.yield..tons.ha.1.", "Sand", "Silt", "Clay", "Soil.pH"),
  soil_climate = c("Grain.yield..tons.ha.1.", "Sand", "Silt", "Clay", "Soil.pH", 
                   "temp1", "temp2", "temp3", "prc1", "prc2", "prc3"),
  extended = c("Grain.yield..tons.ha.1.", "Sand", "Silt", "Clay", "Soil.pH",
               "temp1", "temp2", "temp3", "temp4", "temp5", "temp6",
               "prc1", "prc2", "prc3", "prc4", "prc5", "prc6"),
  comprehensive = c("Grain.yield..tons.ha.1.", "Sand", "Silt", "Clay", "Soil.pH",
                    "temp1", "temp2", "temp3", "temp4", "temp5", "temp6",
                    "prc1", "prc2", "prc3", "prc4", "prc5", "prc6",
                    "AEZ", "Elevation", "N.rate..kg.N.ha.1.", "P.rate..kg.P.ha.1.")
)

# Fast sample size calculation
calc_samples_fast <- function(dataset, var_sets, name) {
  results <- data.frame()
  
  for(set_name in names(var_sets)) {
    vars <- var_sets[[set_name]]
    available_vars <- intersect(vars, names(dataset))
    
    if(length(available_vars) > 1) {
      subset_data <- dataset[, available_vars, drop = FALSE]
      complete_cases <- sum(complete.cases(subset_data))
      percent_complete <- round(complete_cases / nrow(dataset) * 100, 1)
      
      results <- rbind(results, data.frame(
        Dataset = name,
        Variable_Set = set_name,
        Available_Vars = length(available_vars),
        Complete_Cases = complete_cases,
        Percent_Complete = percent_complete
      ))
    }
  }
  return(results)
}

# Calculate for both datasets
na_samples <- calc_samples_fast(north_america, variable_sets, "North America")
global_samples <- calc_samples_fast(global_dataset, variable_sets, "Global")

# Display results
cat("EFFECTIVE SAMPLE SIZE COMPARISON:\n")
cat("Variable Set     | Dataset      | Vars | Complete Cases | % Complete\n")
cat("-----------------|--------------|------|----------------|----------\n")

for(set_name in names(variable_sets)) {
  na_row <- na_samples[na_samples$Variable_Set == set_name, ]
  global_row <- global_samples[global_samples$Variable_Set == set_name, ]
  
  if(nrow(na_row) > 0) {
    cat(sprintf("%-16s | %-12s | %-4d | %-14d | %-9.1f%%\n",
                set_name, "N. America", na_row$Available_Vars,
                na_row$Complete_Cases, na_row$Percent_Complete))
  }
  if(nrow(global_row) > 0) {
    cat(sprintf("%-16s | %-12s | %-4d | %-14d | %-9.1f%%\n",
                "", "Global", global_row$Available_Vars,
                global_row$Complete_Cases, global_row$Percent_Complete))
  }
  cat("\n")
}

# Key insights
extended_na <- na_samples[na_samples$Variable_Set == "extended", ]
extended_global <- global_samples[global_samples$Variable_Set == "extended", ]

if(nrow(extended_na) > 0 && nrow(extended_global) > 0) {
  cat("KEY INSIGHTS (Extended variable set):\n")
  cat("North America:", extended_na$Complete_Cases, "usable observations\n")
  cat("Global:", extended_global$Complete_Cases, "usable observations\n")
  cat("Difference:", extended_global$Complete_Cases - extended_na$Complete_Cases, "observations\n")
  cat("NA as % of Global effective sample:", round(extended_na$Complete_Cases / extended_global$Complete_Cases * 100, 1), "%\n\n")
}
```

## 3. Basic Statistical Summary

```{r basic-stats}
cat("=== BASIC STATISTICAL SUMMARY ===\n")

# Key variables for comparison
key_vars <- c("yield_value", "Sand", "Silt", "Clay", "Soil.pH", "temp1", "prc1")
available_key <- intersect(key_vars, names(clean_data))

cat("Statistical comparison for", length(available_key), "key variables:\n\n")

# Fast statistical comparison
for(var in available_key) {
  if(var %in% names(north_america) && var %in% names(global_dataset)) {
    
    na_vals <- north_america[[var]][!is.na(north_america[[var]])]
    global_vals <- global_dataset[[var]][!is.na(global_dataset[[var]])]
    
    if(length(na_vals) > 10 && length(global_vals) > 10) {
      cat(var, ":\n")
      cat("  North America - N:", length(na_vals), 
          "Mean:", round(mean(na_vals), 3), 
          "SD:", round(sd(na_vals), 3), "\n")
      cat("  Global - N:", length(global_vals), 
          "Mean:", round(mean(global_vals), 3), 
          "SD:", round(sd(global_vals), 3), "\n")
      
      # Simple effect size
      pooled_sd <- sqrt(((length(na_vals)-1)*var(na_vals) + (length(global_vals)-1)*var(global_vals)) / 
                       (length(na_vals) + length(global_vals) - 2))
      cohens_d <- abs(mean(na_vals) - mean(global_vals)) / pooled_sd
      
      effect_size <- ifelse(cohens_d < 0.2, "Negligible",
                           ifelse(cohens_d < 0.5, "Small",
                                 ifelse(cohens_d < 0.8, "Medium", "Large")))
      
      cat("  Effect size:", round(cohens_d, 3), "(", effect_size, ")\n\n")
    }
  }
}
```

## 4. Fast ML Performance Test

```{r fast-ml-test}
cat("=== FAST ML PERFORMANCE TEST ===\n")

# Simple ML test with minimal variables to avoid hanging
simple_ml_vars <- c("yield_value", "Sand", "Silt", "Clay")
available_ml_vars <- intersect(simple_ml_vars, names(clean_data))

if(length(available_ml_vars) >= 3) {
  cat("Testing with variables:", paste(available_ml_vars, collapse = ", "), "\n\n")
  
  # Prepare datasets
  na_ml <- north_america[, available_ml_vars, drop = FALSE]
  na_ml <- na_ml[complete.cases(na_ml), ]
  
  global_ml <- global_dataset[, available_ml_vars, drop = FALSE]
  global_ml <- global_ml[complete.cases(global_ml), ]
  
  # Simple test function
  simple_ml_test <- function(data, name) {
    if(nrow(data) < 50) {
      cat(name, ": Insufficient data\n")
      return(NA)
    }
    
    set.seed(123)
    train_size <- floor(0.8 * nrow(data))
    train_idx <- sample(nrow(data), train_size)
    
    train_data <- data[train_idx, ]
    test_data <- data[-train_idx, ]
    
    # Simple linear model (fast and reliable)
    lm_model <- lm(yield_value ~ ., data = train_data)
    predictions <- predict(lm_model, test_data)
    
    # Calculate R²
    ss_res <- sum((test_data$yield_value - predictions)^2)
    ss_tot <- sum((test_data$yield_value - mean(test_data$yield_value))^2)
    r2 <- 1 - (ss_res / ss_tot)
    
    cat(name, "Results:\n")
    cat("  Training size:", nrow(train_data), "\n")
    cat("  Test size:", nrow(test_data), "\n")
    cat("  R²:", round(r2, 3), "\n\n")
    
    return(r2)
  }
  
  # Run tests
  na_r2 <- simple_ml_test(na_ml, "NORTH AMERICA")
  global_r2 <- simple_ml_test(global_ml, "GLOBAL")
  
  # Compare performance
  if(!is.na(na_r2) && !is.na(global_r2)) {
    cat("PERFORMANCE COMPARISON:\n")
    if(na_r2 > global_r2) {
      advantage <- round((na_r2 - global_r2) / global_r2 * 100, 1)
      cat("✓ North America performs", advantage, "% better\n")
      ml_winner <- "North America"
    } else {
      advantage <- round((global_r2 - na_r2) / na_r2 * 100, 1)
      cat("✓ Global dataset performs", advantage, "% better\n")
      ml_winner <- "Global"
    }
  } else {
    ml_winner <- "Inconclusive"
  }
} else {
  cat("Insufficient variables for ML testing\n")
  ml_winner <- "Inconclusive"
}
```

## 5. Research Interpretability

```{r interpretability}
cat("=== RESEARCH INTERPRETABILITY ASSESSMENT ===\n")

# Simplified interpretability scoring
interpretability_scores <- data.frame(
  Factor = c(
    "Agricultural System Homogeneity",
    "Data Quality and Completeness",
    "Geographic Coherence", 
    "Policy Relevance",
    "Methodological Clarity",
    "Sample Size and Power",
    "Generalizability"
  ),
  
  North_America = c(90, 85, 95, 95, 85, 60, 60),
  Global = c(40, 75, 30, 70, 50, 90, 95),
  
  Weight = c(0.15, 0.20, 0.10, 0.15, 0.15, 0.15, 0.10)
)

# Calculate weighted scores
na_interp_score <- sum(interpretability_scores$North_America * interpretability_scores$Weight)
global_interp_score <- sum(interpretability_scores$Global * interpretability_scores$Weight)

cat("INTERPRETABILITY COMPARISON:\n")
cat("Factor                      | Weight | NA Score | Global | Weighted Diff\n")
cat("----------------------------|--------|----------|--------|---------------\n")

for(i in 1:nrow(interpretability_scores)) {
  row <- interpretability_scores[i, ]
  weighted_diff <- (row$North_America - row$Global) * row$Weight
  
  cat(sprintf("%-27s | %-6.2f | %-8d | %-6d | %-+13.2f\n",
              row$Factor, row$Weight, row$North_America, row$Global, weighted_diff))
}

cat("\nWEIGHTED INTERPRETABILITY SCORES:\n")
cat("North America:", round(na_interp_score, 1), "/100\n")
cat("Global:", round(global_interp_score, 1), "/100\n")

if(na_interp_score > global_interp_score) {
  cat("✓ North America has better research interpretability\n")
  interp_winner <- "North America"
} else {
  cat("✓ Global dataset has better research interpretability\n")
  interp_winner <- "Global"
}
```

## 6. Final Recommendation

```{r final-recommendation}
cat("=== FINAL RECOMMENDATION ===\n\n")

# Simple scoring system
factors <- c("sample_size", "data_quality", "ml_performance", "interpretability")
scores <- data.frame(
  Factor = c("Sample Size", "Data Quality", "ML Performance", "Interpretability"),
  Weight = c(0.25, 0.30, 0.25, 0.20),
  NA_Score = numeric(4),
  Global_Score = numeric(4)
)

# Sample size scoring
if(exists("extended_na") && exists("extended_global")) {
  if(extended_global$Complete_Cases > extended_na$Complete_Cases) {
    scores$NA_Score[1] <- 30
    scores$Global_Score[1] <- 70
  } else {
    scores$NA_Score[1] <- 70
    scores$Global_Score[1] <- 30
  }
}

# Data quality scoring
if(improvement > 2) {
  scores$NA_Score[2] <- 70
  scores$Global_Score[2] <- 30
} else if(improvement < -2) {
  scores$NA_Score[2] <- 30
  scores$Global_Score[2] <- 70
} else {
  scores$NA_Score[2] <- 50
  scores$Global_Score[2] <- 50
}

# ML performance scoring
if(exists("ml_winner")) {
  if(ml_winner == "North America") {
    scores$NA_Score[3] <- 70
    scores$Global_Score[3] <- 30
  } else if(ml_winner == "Global") {
    scores$NA_Score[3] <- 30
    scores$Global_Score[3] <- 70
  } else {
    scores$NA_Score[3] <- 50
    scores$Global_Score[3] <- 50
  }
}

# Interpretability scoring
if(interp_winner == "North America") {
  scores$NA_Score[4] <- 80
  scores$Global_Score[4] <- 20
} else {
  scores$NA_Score[4] <- 20
  scores$Global_Score[4] <- 80
}

# Calculate final weighted scores
na_final <- sum(scores$NA_Score * scores$Weight)
global_final <- sum(scores$Global_Score * scores$Weight)

cat("DECISION MATRIX:\n")
cat("Factor           | Weight | NA Score | Global | Winner\n")
cat("-----------------|--------|----------|--------|----------\n")

for(i in 1:nrow(scores)) {
  winner <- ifelse(scores$NA_Score[i] > scores$Global_Score[i], "N. America", 
                  ifelse(scores$Global_Score[i] > scores$NA_Score[i], "Global", "Tie"))
  
  cat(sprintf("%-16s | %-6.2f | %-8.0f | %-6.0f | %s\n",
              scores$Factor[i], scores$Weight[i], 
              scores$NA_Score[i], scores$Global_Score[i], winner))
}

cat("\nFINAL WEIGHTED SCORES:\n")
cat("North America:", round(na_final, 1), "/100\n")
cat("Global:", round(global_final, 1), "/100\n")

# Final recommendation
if(na_final > global_final) {
  recommendation <- "North America"
  confidence <- round(na_final / (na_final + global_final) * 100, 1)
} else {
  recommendation <- "Global"
  confidence <- round(global_final / (na_final + global_final) * 100, 1)
}

cat("\n🎯 FINAL RECOMMENDATION:", recommendation, "\n")
cat("Confidence Level:", confidence, "%\n")

# Rationale
cat("\nRATIONALE:\n")
if(recommendation == "North America") {
  cat("✓ Better data quality outweighs smaller sample size\n")
  cat("✓ Superior research interpretability\n")
  cat("✓ Reduced confounding from homogeneous systems\n")
  cat("✓ More actionable insights for regional policy\n")
} else {
  cat("✓ Larger sample size provides better statistical power\n")
  cat("✓ Broader generalizability\n")
  cat("✓ Global applicability of results\n")
  cat("✓ Higher publication impact potential\n")
}

cat("\nNEXT STEPS:\n")
cat("1. Implement the recommended approach\n")
cat("2. Document this decision in your methodology\n")
cat("3. Plan validation appropriate to your choice\n")
cat("4. Consider limitations and acknowledge trade-offs\n")
```

## Conclusion

This analysis provides a fast, comprehensive comparison of all variables while avoiding computational bottlenecks. The recommendation is based on:

- **Complete variable inventory** across all categories
- **Effective sample sizes** for realistic modeling scenarios
- **Data quality assessment** with specific improvements quantified
- **Statistical summaries** of key variables
- **ML performance testing** with efficient algorithms
- **Research interpretability** framework

The decision balances sample size vs. data quality trade-offs while considering your research objectives and practical constraints.