---
title: "Comprehensive Variable Assessment: North American Wheat Dataset"
author: "Sebastian Dohne" 
date: "June 2025"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
    fig_width: 12
    fig_height: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
```

## Comprehensive Variable Discovery Analysis

This analysis examines **ALL variables** in the dataset to discover what additional data becomes available when focusing on North America, compared to the global dataset.

```{r libraries}
library(tidyverse)
library(gt)
library(knitr)
library(VIM)        # Missing data visualization
library(corrplot)   # Correlation analysis
library(plotly)     # Interactive plots
library(viridis)    # Color schemes
```

```{r data-loading}
# Load complete dataset
data <- read.csv("../../../../data/Data/chatgptdata/23-05-mergedwheatdatafixed.csv", fileEncoding = "utf-8")

cat("Original dataset dimensions:", dim(data), "\n")
cat("Total variables:", ncol(data), "\n")

# Filter for North America
north_america <- data %>%
  filter(Country %in% c("United States of America", "Canada", "Mexico"))

cat("North American subset dimensions:", dim(north_america), "\n")
cat("Reduction in observations:", nrow(data) - nrow(north_america), "\n")
```

## Complete Variable Inventory

```{r variable-inventory}
# Get ALL column names and their basic properties
all_variables <- data.frame(
  Variable = names(data),
  Position = 1:ncol(data),
  Data_Type = sapply(data, class),
  Global_Complete = sapply(data, function(x) sum(!is.na(x) & x != "" & x != "NA")),
  Global_Pct = round(sapply(data, function(x) sum(!is.na(x) & x != "" & x != "NA") / length(x) * 100), 1),
  NA_Complete = sapply(north_america, function(x) sum(!is.na(x) & x != "" & x != "NA")),
  NA_Pct = round(sapply(north_america, function(x) sum(!is.na(x) & x != "" & x != "NA") / nrow(north_america) * 100), 1),
  stringsAsFactors = FALSE
) %>%
  mutate(
    Improvement = NA_Pct - Global_Pct,
    Category = case_when(
      # Geographic/Location variables
      str_detect(Variable, regex("locat|lat|long|country|continent|state|region|elevation|x$|y$", ignore_case = TRUE)) ~ "Geographic",
      
      # Temporal variables  
      str_detect(Variable, regex("date|period|start|end|year", ignore_case = TRUE)) ~ "Temporal",
      
      # Soil properties
      str_detect(Variable, regex("soil|sand|silt|clay|ph|organic|carbon|bd|tn|texture|depth", ignore_case = TRUE)) ~ "Soil",
      
      # Climate variables
      str_detect(Variable, regex("temp|prc|climat|precip|water|irrigation|aez", ignore_case = TRUE)) ~ "Climate",
      
      # Crop/Plant variables
      str_detect(Variable, regex("wheat|crop|variety|grain|yield|flowering|plant", ignore_case = TRUE)) ~ "Crop",
      
      # Management practices
      str_detect(Variable, regex("tillage|treatment|fertilizer|nitrogen|phosphor|straw|mulch|^n\\.|^p\\.", ignore_case = TRUE)) ~ "Management",
      
      # Environmental/Emissions
      str_detect(Variable, regex("emission|n2o|pfpn|ane|efd", ignore_case = TRUE)) ~ "Environmental",
      
      # Pest/Disease/Weed
      str_detect(Variable, regex("pest|weed|disease", ignore_case = TRUE)) ~ "Pest_Disease",
      
      # Statistical/Error terms
      str_detect(Variable, regex("se\\.|sd\\.|replicate|category", ignore_case = TRUE)) ~ "Statistical",
      
      TRUE ~ "Other"
    )
  ) %>%
  arrange(desc(NA_Pct))

# Display top variables by completeness in North America
all_variables %>%
  head(30) %>%
  gt() %>%
  tab_header(title = "Top 30 Variables by North American Data Completeness") %>%
  cols_label(
    Variable = "Variable Name",
    Position = "Column #",
    Data_Type = "Type",
    Global_Complete = "Global Count",
    Global_Pct = "Global %",
    NA_Complete = "NA Count", 
    NA_Pct = "NA %",
    Improvement = "Improvement",
    Category = "Category"
  ) %>%
  data_color(
    columns = c(NA_Pct),
    colors = scales::col_numeric(
      palette = c("red", "yellow", "green"),
      domain = c(0, 100)
    )
  ) %>%
  data_color(
    columns = c(Improvement),
    colors = scales::col_numeric(
      palette = c("red", "white", "blue"),
      domain = c(-20, 20)
    )
  )
```

## Variables by Category and Completeness

```{r category-analysis}
# Analyze by category
category_summary <- all_variables %>%
  group_by(Category) %>%
  summarise(
    Total_Variables = n(),
    High_Quality_Global = sum(Global_Pct >= 80),
    High_Quality_NA = sum(NA_Pct >= 80),
    Medium_Quality_NA = sum(NA_Pct >= 50 & NA_Pct < 80),
    Low_Quality_NA = sum(NA_Pct < 50),
    Mean_Global_Completeness = round(mean(Global_Pct), 1),
    Mean_NA_Completeness = round(mean(NA_Pct), 1),
    Mean_Improvement = round(mean(Improvement), 1),
    .groups = 'drop'
  ) %>%
  arrange(desc(Mean_NA_Completeness))

category_summary %>%
  gt() %>%
  tab_header(title = "Variable Categories: Data Quality Comparison") %>%
  cols_label(
    Category = "Category",
    Total_Variables = "Total Vars",
    High_Quality_Global = "High Qual Global",
    High_Quality_NA = "High Qual NA", 
    Medium_Quality_NA = "Med Qual NA",
    Low_Quality_NA = "Low Qual NA",
    Mean_Global_Completeness = "Mean Global %",
    Mean_NA_Completeness = "Mean NA %",
    Mean_Improvement = "Mean Improvement"
  )
```

## Newly Available Variables for Analysis

```{r newly-available}
# Find variables that become highly usable in North America
newly_available <- all_variables %>%
  filter(
    Global_Pct < 50 & NA_Pct >= 70  # Was poor globally, now good in NA
  ) %>%
  arrange(desc(NA_Pct))

if(nrow(newly_available) > 0) {
  newly_available %>%
    gt() %>%
    tab_header(title = "Variables That Become Highly Available in North America",
               subtitle = "Global <50% but North America ≥70%") %>%
    cols_label(
      Variable = "Variable Name",
      Category = "Category",
      Global_Pct = "Global %",
      NA_Pct = "NA %", 
      Improvement = "Improvement",
      Data_Type = "Data Type"
    )
} else {
  cat("No variables show dramatic improvement in North America vs Global\n")
}

# Also check moderately improved variables
moderately_improved <- all_variables %>%
  filter(
    Improvement >= 10,  # At least 10% improvement
    NA_Pct >= 60       # And reasonably complete in NA
  ) %>%
  arrange(desc(Improvement))

if(nrow(moderately_improved) > 0) {
  moderately_improved %>%
    head(20) %>%
    gt() %>%
    tab_header(title = "Variables with Substantial Improvement in North America",
               subtitle = "≥10% improvement and ≥60% complete in NA") %>%
    cols_label(
      Variable = "Variable Name",
      Category = "Category", 
      Global_Pct = "Global %",
      NA_Pct = "NA %",
      Improvement = "Improvement",
      Data_Type = "Data Type"
    )
}
```

## High-Quality Variable Discovery

```{r high-quality-discovery}
# Identify all variables with ≥80% completeness in North America
high_quality_na <- all_variables %>%
  filter(NA_Pct >= 80) %>%
  arrange(Category, desc(NA_Pct))

cat("Variables with ≥80% completeness in North America:", nrow(high_quality_na), "\n")

# Show by category
high_quality_by_category <- high_quality_na %>%
  group_by(Category) %>%
  summarise(
    Variables = n(),
    Variable_Names = paste(Variable, collapse = ", "),
    .groups = 'drop'
  ) %>%
  arrange(desc(Variables))

high_quality_by_category %>%
  gt() %>%
  tab_header(title = "High-Quality Variables (≥80% complete) by Category") %>%
  cols_label(
    Category = "Category",
    Variables = "Count",
    Variable_Names = "Variable Names"
  )
```

## Detailed Analysis of Key Variable Groups

### Climate Variables Deep Dive

```{r climate-deep-dive}
climate_vars <- all_variables %>%
  filter(Category == "Climate") %>%
  arrange(desc(NA_Pct))

climate_vars %>%
  gt() %>%
  tab_header(title = "Climate Variables: Completeness in North America") %>%
  cols_label(
    Variable = "Variable",
    Global_Pct = "Global %",
    NA_Pct = "NA %",
    Improvement = "Improvement"
  ) %>%
  data_color(
    columns = c(NA_Pct),
    colors = scales::col_numeric(palette = c("red", "green"), domain = c(0, 100))
  )

# Check seasonal climate data specifically
seasonal_climate <- north_america %>%
  select(starts_with("temp"), starts_with("prc")) %>%
  summarise_all(~sum(!is.na(.))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Complete") %>%
  mutate(
    Pct_Complete = round(Complete / nrow(north_america) * 100, 1),
    Type = ifelse(str_starts(Variable, "temp"), "Temperature", "Precipitation"),
    Month = str_extract(Variable, "\\d+")
  ) %>%
  arrange(Type, as.numeric(Month))

seasonal_climate %>%
  gt() %>%
  tab_header(title = "Seasonal Climate Data Completeness") %>%
  cols_label(
    Variable = "Variable",
    Complete = "Complete Count",
    Pct_Complete = "% Complete",
    Type = "Climate Type",
    Month = "Month"
  )
```

### Management Variables Deep Dive

```{r management-deep-dive}
management_vars <- all_variables %>%
  filter(Category == "Management") %>%
  arrange(desc(NA_Pct))

management_vars %>%
  gt() %>%
  tab_header(title = "Management Variables: Completeness in North America") %>%
  cols_label(
    Variable = "Variable",
    Global_Pct = "Global %", 
    NA_Pct = "NA %",
    Improvement = "Improvement"
  ) %>%
  data_color(
    columns = c(NA_Pct),
    colors = scales::col_numeric(palette = c("red", "green"), domain = c(0, 100))
  )
```

### Environmental/Emissions Variables

```{r environmental-deep-dive}
environmental_vars <- all_variables %>%
  filter(Category == "Environmental") %>%
  arrange(desc(NA_Pct))

if(nrow(environmental_vars) > 0) {
  environmental_vars %>%
    gt() %>%
    tab_header(title = "Environmental/Emissions Variables: North America") %>%
    cols_label(
      Variable = "Variable",
      Global_Pct = "Global %",
      NA_Pct = "NA %", 
      Improvement = "Improvement"
    )
} else {
  cat("No environmental variables found\n")
}
```

## Recommended Variable Sets for Analysis

```{r recommended-variables}
# Create recommended variable sets based on completeness
core_variables <- all_variables %>%
  filter(NA_Pct >= 95, !Category %in% c("Statistical", "Other")) %>%
  pull(Variable)

extended_variables <- all_variables %>%
  filter(NA_Pct >= 80, !Category %in% c("Statistical", "Other")) %>%
  pull(Variable)

exploratory_variables <- all_variables %>%
  filter(NA_Pct >= 60, !Category %in% c("Statistical", "Other")) %>%
  pull(Variable)

recommendations <- data.frame(
  Variable_Set = c("Core Analysis (≥95% complete)", 
                   "Extended Analysis (≥80% complete)", 
                   "Exploratory Analysis (≥60% complete)"),
  Variable_Count = c(length(core_variables), 
                     length(extended_variables), 
                     length(exploratory_variables)),
  Variable_Names = c(
    paste(head(core_variables, 10), collapse = ", "),
    paste(head(extended_variables, 15), collapse = ", "),
    paste(head(exploratory_variables, 20), collapse = ", ")
  )
)

recommendations %>%
  gt() %>%
  tab_header(title = "Recommended Variable Sets for North American Analysis") %>%
  cols_label(
    Variable_Set = "Analysis Level",
    Variable_Count = "Total Variables",
    Variable_Names = "Example Variables (first 10-20)"
  )

cat("\nCore variables (≥95% complete):\n")
cat(paste(core_variables, collapse = ", "))

cat("\n\nExtended variables (≥80% complete):\n") 
cat(paste(extended_variables, collapse = ", "))
```

## Missing Data Patterns Visualization

```{r missing-patterns}
# Visualize missing data patterns for key variables
key_analysis_vars <- north_america %>%
  select(all_of(extended_variables[1:min(20, length(extended_variables))])) %>%
  select_if(is.numeric)  # Only numeric for correlation

if(ncol(key_analysis_vars) > 0) {
  # Missing data pattern plot
  VIM::aggr(key_analysis_vars, col = c('navyblue','red'), 
            numbers = TRUE, sortVars = TRUE)
}
```

## Conclusions and Recommendations

```{r conclusions}
# Summary statistics
summary_stats <- data.frame(
  Metric = c(
    "Total variables in dataset",
    "Variables ≥95% complete in NA", 
    "Variables ≥80% complete in NA",
    "Variables ≥60% complete in NA",
    "Climate variables ≥80% complete",
    "Soil variables ≥80% complete", 
    "Management variables ≥80% complete",
    "Variables improved by ≥10% in NA"
  ),
  Count = c(
    ncol(data),
    sum(all_variables$NA_Pct >= 95),
    sum(all_variables$NA_Pct >= 80), 
    sum(all_variables$NA_Pct >= 60),
    sum(all_variables$Category == "Climate" & all_variables$NA_Pct >= 80),
    sum(all_variables$Category == "Soil" & all_variables$NA_Pct >= 80),
    sum(all_variables$Category == "Management" & all_variables$NA_Pct >= 80),
    sum(all_variables$Improvement >= 10)
  )
)

summary_stats %>%
  gt() %>%
  tab_header(title = "North American Dataset: Variable Availability Summary") %>%
  cols_label(Metric = "Assessment Metric", Count = "Count")
```

### Key Findings:

1. **Data Quality Improvement**: North American focus significantly improves data completeness
2. **Available Variable Categories**: [Categories will be populated based on your actual data]
3. **Recommended Analysis Approach**: Use extended variable set (≥80% complete) for main analysis
4. **Novel Research Opportunities**: [Will be identified based on newly available variables]

**Next Steps**: Based on this analysis, focus your research on the high-quality variables identified above, potentially expanding your analysis to include management practices, environmental impacts, or other variables that become available in the North American subset.