---
title: "Wheat Dataset Exploration and Analysis"
author: "Sebastian Dohne"
date: "June 2025"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
    fig_width: 12
    fig_height: 8
  pdf_document:
    toc: true
    fig_width: 12
    fig_height: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
```

## Project overview

The this is the dataset I have for my final project detailed below titled:  Predicting crop yields in selected region using multivariate statistical and ML approaches

This project aims to predict crop yields in a selected region by leveraging data-driven methods, including multivariate statistical analyses and machine learning (ML) approaches. the dataset includes a combination of environmental, climatic, and socio-economic data to build robust models that predict crop yields with high accuracy. The project will involve tasks such as data preprocessing, covariate selection, spatial and temporal autocorrelation detection, and model tuning. 

Currently, the methods that I am looking at in regards to ML are: Random Forest, XGboost and elastic nets, ANNs and possibly SVRs where deemed appropriate, as these methods can handle categorical and numerical datatypes. From there, my next step was to use more traditional approaches, such as general additive models and non-linear regression to do analyses with the variables deemed most important via ML. 

## Overview

This analysis explores a wheat dataset containing yield information across different countries, locations, and time periods. The primary objectives are:

- Perform comprehensive data cleaning and quality assessment
- Create interactive visualizations to understand geographic and temporal patterns
- Assess data completeness and identify missing value patterns
- Evaluate the dataset's suitability for machine learning applications using quantitative metrics

## Load Required Libraries

```{r libraries}
  library(data.table)    # For fast file reading
  library(tidyverse)     # Data manipulation
  library(knitr)         # Tables
  library(ggplot2)       # Static plots only
  library(scales)        # For formatting
  library(leaflet)
  library(gt)
  library(kableExtra)
  library(plotly)  # Add this to your libraries section
  library(terra)     # For raster data handling (modern replacement for raster package)
  library(leaflet)   # Already in your code
  library(leafem)    # Optional: provides addGeotiff() for better TIF handling
  library(rasterVis)
  library(viridis)

# Parallel processing
library(furrr)        # Parallel mapping functions
```

## Data Loading and Initial Exploration

```{r data-loading}
# Load wheat data 
# Note: fileEncoding handles special characters in country/location names
# Update this path to match where your CSV file is located:
data <- read.csv("../../../../data/Data/explorationdatasets/AllbutNA.csv", fileEncoding = "utf-8")

# Loading AEZ raster data
aez_raster <- terra::rast("../../../../data/MSc - Sebastian/gbcl-wheat-msc/data-raw/otherdata/aez_v9v2red_5m_CRUTS32_Hist_8110_100_avg.tif")

```

### Initial Dataset inspection

```{r}
cat("Dataset dimensions:", dim(data), "\n")
cat("Number of countries:", length(unique(data$Country)), "\n")
cat("Time period covered:", range(data$Observation.period, na.rm = TRUE), "\n")
```


```{r column-names}
# Display all column names for reference
names(data)
```

## Data Cleaning and Preprocessing

### Column Selection and Type Conversion

```{r data-cleaning}
# Select key variables and rename for easier handling
yield <- data %>%
  select(Country, 
         Observation.period, 
         Location, 
         Continent, 
         State.Region.County.Province, 
         longitude = Conversion.for.longitude,    # Rename for clarity
         latitude = Conversion.for.latitude,      # Rename for clarity
         yield_value = Grain.yield..tons.ha.1.) %>%  # Main outcome variable
  mutate(
    # Convert character columns to appropriate numeric types
    longitude = as.numeric(longitude),
    latitude = as.numeric(latitude),
    yield_value = as.numeric(yield_value),
    Observation.period = as.numeric(Observation.period)
  ) %>%  
  filter(
    # Remove problematic observations
    !is.na(yield_value),                    # Remove missing yield values
    !is.na(longitude),                      # Remove missing coordinates
    !is.na(latitude),
    yield_value <= 30,                      # Remove unrealistic yield outliers (>30 tons/ha)
    nchar(as.character(Observation.period)) == 4  # Keep only 4-digit years
  )
cat("Yield range:", range(yield$yield_value), "tons/ha\n")
```

## Geographic Visualization

### Interactive Leaflet Map

```{r}
plot(aez_raster)  # Check if it renders properly
```


```{r leaflet-map}
# Ensure CRS is defined and reproject if needed
if (is.na(crs(aez_raster))) crs(aez_raster) <- "EPSG:4326"
if (crs(aez_raster) != "EPSG:4326") aez_raster <- project(aez_raster, "EPSG:4326")

# Downsample if too large
if (ncell(aez_raster) > 1e6) aez_raster <- aggregate(aez_raster, fact = 10)

# Convert raster to dataframe for ggplot
r_df <- as.data.frame(aez_raster, xy = TRUE)
colnames(r_df)[3] <- "zone_value"

# Convert yield data to sf
yield_sf <- st_as_sf(yield, coords = c("longitude", "latitude"), crs = 4326)

# Plot
ggplot() +
  geom_raster(data = r_df, aes(x = x, y = y, fill = zone_value)) +
  scale_fill_viridis(name = "Agroecological Zone", na.value = NA) +
  geom_sf(data = yield_sf, aes(color = yield_value), size = 2, alpha = 0.8) +
  scale_color_viridis(name = "Yield (tons/ha)", option = "inferno") +
  theme_minimal() +
  labs(title = "Agroecological Zones and Wheat Yield Points") +
  coord_sf()
```

**Map Features:**
- **Color coding:** Yellow to red gradient representing yield intensity
- **Interactive popups:** Click markers for detailed information
- **Legend:** Helps interpret yield values across locations

## Temporal Analysis

### Distribution of Observations Over Time

```{r temporal-analysis}
# Histogram showing data collection patterns over years
ggplot(yield, aes(x = Observation.period)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Distribution of Wheat Yield Observations Over Time", 
       x = "Year", 
       y = "Number of Observations") +
  theme(plot.title = element_text(size = 14, face = "bold"))
```

### Yearly Summary Statistics

```{r yearly-summary}
# Summary of observations and yield by year
yearly_summary <- yield %>%
  group_by(Observation.period) %>%
  summarise(
    n_observations = n(),
    mean_yield = round(mean(yield_value), 2),
    median_yield = round(median(yield_value), 2),
    n_countries = n_distinct(Country),
    .groups = 'drop'
  ) %>%
  arrange(desc(Observation.period))

# Display recent years
head(yearly_summary, 10) %>%
  kable(caption = "Recent Years: Observations and Yield Statistics")
```

## Geographic Distribution Analysis

### Observations by Country and Continent

```{r country-distribution}
# Bar plot showing observations by country, colored by continent
ggplot(yield, aes(x = reorder(Country, -table(Country)[Country]), fill = Continent)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Number of Wheat Yield Observations by Country and Continent", 
       x = "Country", 
       y = "Number of Observations") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 14, face = "bold")) +
  scale_fill_viridis_d(option = "plasma")
```

### Country-Continent Summary Table

```{r country-table}
# Create organized summary table by country and continent
country_continent_table <- yield %>%
  group_by(Country, Continent) %>%
  summarise(
    count = n(),
    mean_yield = round(mean(yield_value), 2),
    .groups = 'drop'
  ) %>%
  arrange(desc(count))

# Display as formatted table
country_continent_table %>%
  head(15) %>%
  gt() %>%
  tab_header(title = "Top 15 Countries by Number of Observations") %>%
  cols_label(
    Country = "Country",
    Continent = "Continent", 
    count = "Observations",
    mean_yield = "Mean Yield (tons/ha)"
  ) %>%
  fmt_number(columns = mean_yield, decimals = 2)
```

## Missing Data Analysis

### Missing Data Heatmap by Country

```{r missing-data-prep}
# Prepare data for missing value heatmap
missing_heatmap_data <- data %>%
  group_by(Country) %>%
  summarise(across(everything(), ~mean(is.na(.)) * 100), .groups = 'drop') %>%
  pivot_longer(cols = -Country, 
               names_to = "Variable", 
               values_to = "Percent_Missing")
```

```{r missing-heatmap}
# Create heatmap visualization
ggplot(missing_heatmap_data, aes(x = Variable, y = Country, fill = Percent_Missing)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red", name = "% Missing") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 14, face = "bold")) +
  labs(title = "Missing Data Heatmap by Country", 
       x = "Variable", 
       y = "Country")
```

## Interactive heatmap by country

```{r}
# Percentage missing for all variables
pct_missing_all <- data %>%
  group_by(Country) %>%
  summarise(
    Total_Obs = n(),
    across(everything(), ~round(mean(is.na(.)) * 100, 2), .names = "Pct_NA_{.col}")
  ) %>%
  arrange(desc(Total_Obs))

# First, create the heatmap_data
heatmap_data <- pct_missing_all %>%
  select(-Total_Obs) %>%
  pivot_longer(cols = -Country, 
               names_to = "Variable", 
               values_to = "Pct_Missing",
               names_prefix = "Pct_NA_")

# Create interactive heatmap
p <- plot_ly(
  data = heatmap_data,
  x = ~Variable,
  y = ~Country,
  z = ~Pct_Missing,
  type = "heatmap",
  colorscale = "RdYlBu",
  reversescale = TRUE,
  hovertemplate = "Country: %{y}<br>Variable: %{x}<br>Missing: %{z}%<extra></extra>"
) %>%
  layout(
    title = "Missing Data Heatmap",
    xaxis = list(title = "Variable"),
    yaxis = list(title = "Country")
  )

p
```

## Variable Categories Overview

### Data Type and Category Breakdown

```{r variable-categories}
# Categorize variables by type
location_time_vars <- c("id", "Data.ID", "Location", "State.Region.County.Province", 
                       "Country", "Continent", "Latitude..N.S.", "Longitude..E.W.", 
                       "Conversion.for.latitude", "Conversion.for.longitude", 
                       "Location.source", "Observation.period", "X", "Y", "unit_code")

crop_management_vars <- c("Wheat.Type", "Crop.variety", "Tillage.type", "Planting.date", 
                         "Harvesting.date", "Flowering.stage", "Treatment", "Treatment.type", 
                         "Water.regime", "Irrigation..mm.", "N.type", "N.rate..kg.N.ha.1.", 
                         "N.fertilizer.management", "P.type", "P.rate..kg.P.ha.1.", 
                         "Straw.return", "Plastic.film.mulching", "Planting.date.1", 
                         "Harvesting.date.1", "rainfed_start_date", "rainfed_end_date", 
                         "irr_start_date", "irr_end_date", "start_date", "end_date")

climate_vars <- c("Climate", "Mean.annual.temperature..Â.C.", "Mean.annual.precipitation..mm.", 
                 "temp1", "temp2", "temp3", "temp4", "temp5", "temp6", "temp7", "temp8", "temp9",
                 "prc1", "prc2", "prc3", "prc4", "prc5", "prc6", "prc7", "prc8", "prc9", 
                 "pr_irrigated", "AEZ", "Elevation")

soil_vars <- c("Soil.type", "Soil.depth..cm.", "Sand", "Silt", "Clay", "Soil.texture", 
              "Soil.organic.carbon..g.C.kg.1.", "Soil.organic.carbon....", "TN..g.N.kg.1.", 
              "C.N.ratio", "Soil.pH", "BD..g.cm.3.", "Soil_N")

yield_performance_vars <- c("Grain.yield..tons.ha.1.", "SE...22", "PFPN..kg.kg..1.", 
                           "ANE..kg.kg.1.", "Replicates")

environmental_vars <- c("Cumulative.N2O.fluxes..kg.N.ha.1.", "SE...56", "SD...57", 
                       "Yield.scaled.N2O.emission..g.N.Mg.1.", "SE...59", "SD...60", 
                       "EFd....", "Emissions..yes.no.")

pest_weed_vars <- grep("Pest|weed|Main.weed", names(data), value = TRUE, ignore.case = TRUE)

# Create summary table
category_summary <- data.frame(
  Category = c("Location & Time", "Crop & Management", "Climate", "Soil Properties", 
               "Yield & Performance", "Environmental Impact", "Pest & Weed"),
  Variables = c(length(location_time_vars), length(crop_management_vars), 
                length(climate_vars), length(soil_vars), length(yield_performance_vars),
                length(environmental_vars), length(pest_weed_vars)),
  Data_Types = c("Mixed", "Mixed", "Numeric", "Mixed", "Numeric", "Numeric", "Mixed")
)

category_summary %>%
  gt() %>%
  tab_header(title = "Dataset Structure: Variables by Category") %>%
  cols_label(
    Category = "Variable Category",
    Variables = "Number of Variables",
    Data_Types = "Primary Data Types"
  )

# Show data types breakdown
data_types <- data %>%
  summarise_all(class) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Type") %>%
  count(Type, name = "Count") %>%
  arrange(desc(Count))

data_types %>%
  gt() %>%
  tab_header(title = "Data Types Distribution") %>%
  cols_label(Type = "Data Type", Count = "Number of Variables")
```

## Data Coverage Summary

### Completeness by Variable Category

```{r data-coverage}
# Function to calculate completeness for variable groups
calculate_completeness <- function(data, var_list, category_name) {
  if(length(var_list) > 0) {
    available_vars <- intersect(var_list, names(data))
    if(length(available_vars) > 0) {
      complete_any <- data %>%
        select(all_of(available_vars)) %>%
        mutate(has_any_data = rowSums(!is.na(.)) > 0) %>%
        pull(has_any_data) %>%
        sum()
      
      complete_all <- data %>%
        select(all_of(available_vars)) %>%
        complete.cases() %>%
        sum()
      
      return(data.frame(
        Category = category_name,
        Variables_Available = length(available_vars),
        Obs_With_Any_Data = complete_any,
        Obs_With_Complete_Data = complete_all,
        Pct_Any_Data = round(complete_any / nrow(data) * 100, 1),
        Pct_Complete_Data = round(complete_all / nrow(data) * 100, 1)
      ))
    }
  }
  return(NULL)
}

# Calculate completeness for each category
coverage_summary <- bind_rows(
  calculate_completeness(data, climate_vars, "Climate"),
  calculate_completeness(data, soil_vars, "Soil Properties"),
  calculate_completeness(data, crop_management_vars, "Crop & Management"),
  calculate_completeness(data, yield_performance_vars, "Yield & Performance"),
  calculate_completeness(data, environmental_vars, "Environmental Impact")
)

coverage_summary %>%
  gt() %>%
  tab_header(title = "Data Coverage by Category") %>%
  cols_label(
    Category = "Variable Category",
    Variables_Available = "Variables",
    Obs_With_Any_Data = "Obs. w/ Any Data",
    Obs_With_Complete_Data = "Obs. w/ Complete Data",
    Pct_Any_Data = "% Any Data",
    Pct_Complete_Data = "% Complete"
  )
```

## Variable Completeness Dashboard

### Comprehensive Variable Assessment

```{r completeness-dashboard}
# Create comprehensive variable summary
var_summary <- data %>%
  summarise_all(list(
    count = ~sum(!is.na(.)),
    pct_complete = ~round(sum(!is.na(.)) / n() * 100, 1),
    min_val = ~ifelse(is.numeric(.), round(min(., na.rm = TRUE), 2), NA),
    max_val = ~ifelse(is.numeric(.), round(max(., na.rm = TRUE), 2), NA),
    n_unique = ~n_distinct(., na.rm = TRUE)
  )) %>%
  pivot_longer(everything(), names_to = "variable_stat", values_to = "value") %>%
  separate(variable_stat, into = c("variable", "statistic"), sep = "_(?=count$|pct_complete$|min_val$|max_val$|n_unique$)") %>%
  pivot_wider(names_from = statistic, values_from = value) %>%
  arrange(desc(pct_complete))

# Display all variables by completeness
var_summary %>%
  gt() %>%
  tab_header(title = "All Variables by Data Completeness") %>%
  cols_label(
    variable = "Variable",
    count = "Count",
    pct_complete = "% Complete",
    min_val = "Min Value",
    max_val = "Max Value", 
    n_unique = "Unique Values"
  ) %>%
  fmt_number(columns = c(pct_complete), decimals = 1)
```

### Completeness Visualization

```{r completeness-viz}
# Completeness distribution
ggplot(var_summary, aes(x = pct_complete)) +
  geom_histogram(bins = 20, fill = "steelblue", alpha = 0.7, color = "white") +
  theme_minimal() +
  labs(title = "Distribution of Variable Completeness",
       x = "Percentage Complete",
       y = "Number of Variables") +
  geom_vline(xintercept = 50, color = "red", linetype = "dashed", size = 1) +
  annotate("text", x = 55, y = max(table(cut(var_summary$pct_complete, 20))) * 0.8, 
           label = "50% threshold", color = "red")
```

## Data Source and Quality Indicators

### ML Readiness Assessment

```{r ml-readiness}
# Assess dataset readiness for ML
ml_readiness <- data.frame(
  Assessment_Category = c(
    "Sample Size",
    "Geographic Coverage", 
    "Temporal Coverage",
    "Target Variable (Yield)",
    "Climate Variables",
    "Soil Variables", 
    "Management Variables",
    "Overall ML Readiness"
  ),
  
  Status = c(
    ifelse(nrow(data) >= 100, "✓ Good", "⚠ Limited"),
    ifelse(length(unique(data$Country)) >= 5, "✓ Good", "⚠ Limited"),
    ifelse(max(data$Observation.period, na.rm = TRUE) - min(data$Observation.period, na.rm = TRUE) >= 10, "✓ Good", "⚠ Limited"),
    ifelse(sum(!is.na(data$Grain.yield..tons.ha.1.)) / nrow(data) >= 0.8, "✓ Good", "⚠ Limited"),
    ifelse(sum(!is.na(data$Mean.annual.temperature..Â.C.)) / nrow(data) >= 0.5, "✓ Good", "⚠ Limited"),
    ifelse(sum(!is.na(data$Soil.pH)) / nrow(data) >= 0.3, "✓ Good", "⚠ Limited"),
    ifelse(sum(!is.na(data$N.rate..kg.N.ha.1.)) / nrow(data) >= 0.3, "✓ Good", "⚠ Limited"),
    "Ready for Analysis"
  ),
  
  Details = c(
    paste(nrow(data), "observations"),
    paste(length(unique(data$Country)), "countries"),
    paste(max(data$Observation.period, na.rm = TRUE) - min(data$Observation.period, na.rm = TRUE), "year span"),
    paste(round(sum(!is.na(data$Grain.yield..tons.ha.1.)) / nrow(data) * 100, 1), "% complete"),
    paste(round(sum(!is.na(data$Mean.annual.temperature..Â.C.)) / nrow(data) * 100, 1), "% complete"),
    paste(round(sum(!is.na(data$Soil.pH)) / nrow(data) * 100, 1), "% complete"),
    paste(round(sum(!is.na(data$N.rate..kg.N.ha.1.)) / nrow(data) * 100, 1), "% complete"),
    "Proceed with feature selection"
  )
)

ml_readiness %>%
  gt() %>%
  tab_header(title = "Machine Learning Readiness Assessment") %>%
  cols_label(
    Assessment_Category = "Category",
    Status = "Status",
    Details = "Details"
  )
```